---
title: "Appendices"
author: "Ben Abraham   "
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, include = FALSE}
# Load any packages you may need here
library(car)
library(emmeans)
```


```{r readData, include = FALSE}
income <- read.csv("cancer_mortality_rates.csv", stringsAsFactors = TRUE)
income$medianIncome <- income$medianIncome / 1000
```


***

#### Figure 1: Scatterplot Matrix of Data

```{r figure1, echo=FALSE}
plot(~ medianIncome + pctBach + pctUnemployed + avgHouse, data = income)
```

***

#### Figure 2: Histogram of Median Income

```{r figure2, echo=FALSE}
hist(income$medianIncome,
     main = "Histogram of Median Income",
     xlab = "Body Mass (in grams)",
     col = "deepskyblue")
```

***

#### Figure 3: Histogram of the Proportion of Residents with a Bachelor's Degree

```{r figure3, echo=FALSE}
hist(income$pctBach,
     main = "Histogram of the Proportion of Residents with a Bachelor's Degree",
     xlab = "Proportion of Residents with a Bachelor's Degree",
     col = "darkslategray3")
```

***

#### Figure 4: Histogram of Unemployment Rate

```{r figure4, echo=FALSE}
hist(income$pctUnemployed,
     main = "Histogram of Unemployment Rate",
     xlab = "Unemployment Rate",
     col = "darkseagreen2")
```

***

#### Figure 5: Histogram of Average Household Size

```{r figure5, echo=FALSE}
hist(income$avgHouse,
     main = "Histogram of Average Household Size",
     xlab = "Average Household Size",
     col = "indianred2")
```

***

#### Figure 6: Boxplot of Region

```{r figure6, echo = FALSE}
plot(medianIncome ~ region, data = income, 
     xlab = "Region", ylab = "Median Income (In thousands of dollars)", 
     main = "Median Income by Region", 
     col = c("darkkhaki", "dodgerblue", "darkorange3", "firebrick4", "lightblue1"))
```

***

#### Figure 7: Initial Model Summary

```{r figure 7, echo = FALSE}
model_version0 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + region, data = income)
summary(model_version0)
```

***

#### Figure 8: Distribution of Population Variable Before Log Transformation


```{r figure8, echo = FALSE}
hist(income$population,
     main = "Histogram of Population without Logarithmic Transformation",
     xlab = "Population",
     col = "lightblue")
```

***

#### Figure 9: Distribution of Population Variable After Logarithmic Transformation


```{r figure9, echo = FALSE}
hist(log(income$population),
     main = "Histogram of Population with Logarithmic Transformation",
     xlab = "log(population)",
     col = "lightblue")
```

***

#### Figure 10: Scatterplot of Poverty Rate and Median Income


```{r figure10, echo = FALSE}
plot(medianIncome ~ pctPoverty,
     data = income,
     xlab = "Poverty Rate",
     ylab = "Median Income",
     main = "Median Income vs Poverty Rate")
```

***

#### Figure 11: Scatterplot of Median Income and log(population) by Region


```{r figure11, echo = FALSE}
plot(medianIncome ~ log(population),
     data = income,
     col = c("blue", "orange", "gray40", "red", "lightblue")[region],
     pch = c(1, 2, 3, 4, 5)[region],
     xlab = "log(population)",
     ylab = "Median Income",
     main = "Median Income vs log(population) by Region")

legend(x = 14.5, y = 99,
       legend = levels(income$region),
       col = c("blue", "orange", "gray40", "red", "lightblue"),
       pch = c(1, 2, 3, 4, 5))

```

***

#### Figure 12: Summary Output with Interaction Term on Average House Size


```{r figure12, echo = FALSE}
model_interaction <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + log(population) * region, data = income)
summary(model_interaction)
```

***

#### Figure 13: Scatterplot Matrix of Strong Linear Relationships (pctHS, pctEmployed, pctMarried, and pctMarriedHouse)

```{r figure13, echo=FALSE}
plot(~ medianIncome + pctHS + pctEmployed + pctMarried + pctMarriedHouse, data = income)
```


***

#### Figure 14: Scatterplot Matrix of Strong Linear Relationships (pctPrivateHC, pctPublicHC, and pctEmployerHC)

```{r figure14, echo=FALSE}
plot(~ medianIncome + pctPrivateHC + pctPublicHC + pctEmployerHC, data = income)
```

***

#### Figure 15: Scatterplot of Median Income vs Population with a logarithmic transformation

```{r figure15, echo=FALSE}
plot(medianIncome ~ log(population), data = income,
     main = "Median Income vs log(population)",
     xlab = "log(population)",
     ylab = "Median Income")
```

***

#### Figure 16: Model Output for Initial Model with pctHS as additional predictor

```{r figure16, echo=FALSE}
model_version2 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctHS + region, data = income)
summary(model_version2)
```


***

#### Figure 17: Model Output for Initial Model with pctPublicHC as additional predictor

```{r figure17, echo=FALSE}
model_version3 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctPublicHC + region, data = income)
summary(model_version3)
```

***

#### Figure 18: Model Output for Initial Model with pctPrivateHC as additional predictor

```{r figure18, echo=FALSE}
model_version4 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctPrivateHC + region, data = income)
summary(model_version4)
```

***

#### Figure 19: Model Output for Initial Model with pctEmployerHC as additional predictor

```{r figure19, echo=FALSE}
model_version5 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctEmployerHC + region, data = income)
summary(model_version5)
```

***

#### Figure 20: Model Output for Initial Model with pctMarried as additional predictor

```{r figure20, echo=FALSE}
model_version6 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctMarried + region, data = income)
summary(model_version6)
```


***

#### Figure 21: Model Output for Initial Model with pctPoverty as additional predictor

```{r figure21, echo=FALSE}
model_version7 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctPoverty + I(pctPoverty^2)+ region, data = income)
summary(model_version7)
```



***

#### Figure 22: Model Output for Initial Model with pctMarriedHouse as additional predictor

```{r figure22, echo=FALSE}
model_version8 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctMarriedHouse + region, data = income)
summary(model_version8)
```

***

#### Figure 23: Model Output for Initial Model with log(population) as additional predictor

```{r figure23, echo=FALSE}
model_version9 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + log(population) + region, data = income)
summary(model_version9)
```



***

#### Figure 24: Model Output of Model with Added Variables

```{r figure24, echo=FALSE}
model_final_v1 <- lm(medianIncome ~ pctBach + pctUnemployed + pctPublicHC + pctPrivateHC + pctEmployerHC + pctMarried + pctMarriedHouse + avgHouse + pctPoverty + I(pctPoverty^2) + log(population) + region, data = income)
summary(model_final_v1)
```



***

#### Figure 25: Comparing models with a quadratic term and one without

```{r figure25, echo=FALSE}
model_without_quadratic <- lm(medianIncome ~ pctPoverty, data = income)
model_with_quadratic <- lm(medianIncome ~ pctPoverty + I(pctPoverty^2), data = income)
model_with_third_power <- lm(medianIncome ~ pctPoverty + I(pctPoverty^2) +I(pctPoverty^3), data = income)


quadData <- data.frame(pctPoverty = seq(min(income$pctPoverty), max(income$pctPoverty), by = 0.01))

quadData$Fitted <- predict(model_without_quadratic, newdata = quadData)
quadData$Fitted1 <- predict(model_with_quadratic, newdata = quadData)
quadData$Fitted2 <- predict(model_with_third_power, newdata = quadData)


plot(income$medianIncome ~ income$pctPoverty,
     main = "Median Income vs Poverty by Exponents",
     ylab = "Median Income",
     xlab = "Poverty Rate")


lines(quadData$pctPoverty, quadData$Fitted, col = "red")
lines(quadData$pctPoverty, quadData$Fitted1, col = "blue")
lines(quadData$pctPoverty, quadData$Fitted2, col = "green")

legend(30, 99, legend=c("First Power", "Second Power", "Third Power"),
       col=c("red", "blue", "green"), lty=1, cex=1)

```

***

#### Figure 26: Plot of Residuals for Model of the Second Power

```{r figure26, echo=FALSE}
plot(model_final_v1$fitted.values, model_final_v1$residuals,
     main = "Residuals versus Fitted Values Plot",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```


***

#### Figure 27: Model with third power exponent

```{r figure27, echo=FALSE}
model_final_v2 <- lm(medianIncome ~ pctBach + pctUnemployed + pctPublicHC + pctPrivateHC + pctEmployerHC + pctMarried + pctMarriedHouse + avgHouse + pctPoverty + I(pctPoverty^2) + I(pctPoverty^3) + log(population) + region, data = income)
summary(model_final_v2)
```




***

#### Figure 28: Residual Plot for Final Model

```{r figure28, echo=FALSE}
plot(model_final_v2$fitted.values, model_final_v2$residuals,
     main = "Residuals versus Fitted Values Plot",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```


***

#### Figure 29: QQ Plot for Final Model

```{r figure29, echo=FALSE}
qqnorm(model_final_v2$residuals, main = "QQ Plot of the Residuals")
qqline(model_final_v2$residuals)
```



***

#### Figure 30: VIF Output for Final Model

```{r figure30, echo=FALSE}
vif(model_final_v2)
```

***

#### Figure 31: Expected Change in Response for a one Standard Deviation Increase in each variable

```{r figure31, echo=FALSE}
sd(income$pctBach) * model_final_v2$coefficients[2]
sd(income$pctUnemployed) * model_final_v2$coefficients[3]
sd(income$pctPublicHC) *model_final_v2$coefficients[4]
sd(income$pctPrivateHC) * model_final_v2$coefficients[5]
sd(income$pctEmployerHC) * model_final_v2$coefficients[6]
sd(income$pctMarried) * model_final_v2$coefficients[7]
sd(income$pctMarriedHouse) * model_final_v2$coefficients[8]
sd(income$avgHouse) * model_final_v2$coefficients[9]
sd(log(income$population)) * model_final_v2$coefficients[13]
```

***

#### Figure 32: Model Output without pctMarried

```{r figure32, echo=FALSE}
model_final_v3 <- lm(medianIncome ~ pctBach + pctUnemployed + pctPublicHC + pctPrivateHC + pctEmployerHC + pctMarriedHouse + avgHouse + pctPoverty + I(pctPoverty^2) + + I(pctPoverty^3) + log(population) + region, data = income)
summary(model_final_v3)
```









