---
title: "STATS 401 - Final Project"
author: "Ben Abraham"
date: "March 20th, 2022"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
```

## Cancer Mortality Rates Dataset
> This dataset contains data pertaining to Cancer Mortality Rates from a random sample of 570 counties throughout the United States, and contains the variables listed below:

* cmRate: The number of cancer mortalities per 100 thousand people
* population: The population of the county
* medianIncome: The median income of the county
* binnedIncome: The median income of the county binned by decile
* pctPoverty: The percent of the county living in poverty
* avgHouse: The average household size in terms of the number of residents
* medianAge: The median age of county residents
* medianAgeM: The median age of male county residents
* medianAgeF: The median age of female county residents
* pctHS: The percent of county residents (ages 25+) with a high school diploma
* pctBach: The percent of county residents (ages 25+) with a bachelor's degree
* pctEmployed: The percent of county residents (ages 16+) who are employed
* pctUnemployed: The percent of county residents (ages 16+) who are unemployed
* pctPrivateHC: The percent of county residents with private health coverage
* pctEmployerHC: The percent of county residents with employer-provided health coverage
* pctPublicHC: The percent of county residents with government-provided health coverage
* pctWhite: The percent of county residents who identify as White
* pctBlack: The percent of county residents who identify as Black
* pctAsian: The percent of county residents who identify as Asian
* pctOther: The percent of county residents who identify as a race which is not White, Black, or Asian
* pctMarried: The percent of county residents who are married
* pctMarriedHouse: The percent of married households
* birthRate: The number of live births relative to number of women in county
* region: The region of the county
* state: The state of the county
* county: The name of the county

```{r dataset}
rates <- read.csv("cancer_mortality_rates.csv", stringsAsFactors = TRUE)
rates$medianIncome <- rates$medianIncome / 1000
```


```{r}
head(rates)
```
## Initial Model and Data Exploration

### Variables
- **Response:** Median Income
- **Quantitative Predictor #1:** Percent of Residents with Bachelor's Degree
- **Quantitative Predictor #2:** Unemployment Rate
- **Quantitative Predictor #3:** Average Household Size
- **Categorical Predictor:** Region

```{r numericalSummaries}
# Replace this text with your code
summary(rates)
```

```{r histograms}
# Replace this text with your code
hist(rates$medianIncome,
     main = "Histogram of Median Income (in thousands of dollars)",
     xlab = "Median Income",
     col = "deepskyblue")
hist(rates$pctBach,
     main = "Histogram of Proportions of Residents with Bachelors Degree",
     xlab = "Proportions of Residents with Bachelors Degree",
     col = "darkslategray3")
hist(rates$pctUnemployed,
     main = "Histogram of Unemployment Percent",
     xlab = "Percent of County Residents (Ages 16+) who are Unemployed",
     col = "darkseagreen2")
hist(rates$avgHouse,
     main = "Histogram of Average Household Size",
     xlab = "Average Household Size",
     col = "indianred2")
```

```{r scatterplotMatrix}
# Replace this text with your code
plot( ~ medianIncome + pctBach + pctUnemployed + avgHouse, data = rates)
```
```{r sideBySideBoxplot}
# Replace this text with your code
plot(medianIncome ~ region, data = rates, 
     xlab = "Region", ylab = "Median Income (In thousands of dollars)", 
     main = "Median Income by Region", 
     col = c("darkkhaki", "dodgerblue", "darkorange3", "firebrick4", "lightblue1"))
        
```
```{r intitialModel}
# Replace this text with your code
model <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + region, data = rates)
summary(model)
```




### Scatterplots

```{r}
plot(rates[, c(3, 1, 2, 5, 6)])
```
```{r}
plot(medianIncome ~ log(population),
     data = rates,
     col = c("blue", "orange", "gray40", "red", "lightblue")[region],
     pch = c(1, 2, 3, 4, 5)[region],
     xlab = "log(population)",
     ylab = "Median Income",
     main = "Median Income vs log(population) by Region")

legend(x = 14.5, y = 99,
       legend = levels(rates$region),
       col = c("blue", "orange", "gray40", "red", "lightblue"),
       pch = c(1, 2, 3, 4, 5))
```


```{r}
plot(rates[, c(3, 7, 8, 9)])
```


```{r}
plot(rates[, c(3, 10, 11, 12, 13)])
```
```{r}
plot(rates[, c(3, 14, 15, 16)])
```


```{r}
plot(rates[, c(3, 17, 18, 19, 20)])
```

```{r}
plot(rates[, c(3, 21, 22, 23)])
```

## Final Model Development

### Individual Models

```{r}
summary(model)
model_version2 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctHS + region, data = rates)
summary(model_version2)
```



```{r}
summary(model)
model_version3 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctEmployed + region, data = rates)
summary(model_version3)
```


```{r}
summary(model)
model_version4 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctEmployerHC + region, data = rates)
model_version5 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctPublicHC + region, data = rates)
model_version6 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctPrivateHC + region, data = rates)
summary(model_version4)
summary(model_version5)
summary(model_version6)
```




```{r}
summary(model)
model_version7 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctMarried + region, data = rates)
summary(model_version7)
```


```{r}
summary(model)
model_version8 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctMarriedHouse + region, data = rates)
summary(model_version8)
```



```{r}
summary(model)
model_version9 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctPoverty + region, data = rates)
summary(model_version9)
model_version10 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + pctPoverty + I(pctPoverty^2) + region, data = rates)
summary(model_version10)
```
```{r}
summary(model)
model_version11 <- lm(medianIncome ~ pctBach + pctUnemployed + avgHouse + log(population) + region, data = rates)
summary(model_version11)
```


### Combined Models

```{r}
summary(model)
model_version12 <- lm(medianIncome ~ pctBach + pctUnemployed + pctPublicHC + pctPrivateHC + pctEmployerHC + pctMarried + pctMarriedHouse + avgHouse + pctPoverty + I(pctPoverty^2) + log(population) + region, data = rates)
summary(model_version12)

```

```{r}
plot(model_version12$fitted.values, model_version12$residuals,
     main = "Residuals versus Fitted Values Plot",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
qqnorm(model_version12$residuals, main = "QQ Plot of the Residuals")
qqline(model_version12$residuals)
```

```{r}
model_version13 <- lm(log(medianIncome) ~ pctBach + pctUnemployed + pctPublicHC + pctPrivateHC + pctEmployerHC + pctMarried + pctMarriedHouse + avgHouse + pctPoverty + log(population) + region, data = rates)
summary(model_version12)
summary(model_version13)
```

```{r}
plot(model_version13$fitted.values, model_version13$residuals,
     main = "Residuals versus Fitted Values Plot",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```



```{r}
summary(model_version12)
model_version14 <- lm(medianIncome ~ pctBach + pctUnemployed + pctPublicHC + pctPrivateHC + pctEmployerHC + pctMarried + pctMarriedHouse + avgHouse + pctPoverty + I(pctPoverty^2) + I(pctPoverty^3) + log(population) + region, data = rates)
summary(model_version14)
plot(model_version14$fitted.values, model_version14$residuals,
     main = "Residuals versus Fitted Values Plot",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

sd(rates$pctBach) * model_version14$coefficients[2]
sd(rates$pctUnemployed) * model_version14$coefficients[3]
sd(rates$pctPublicHC) * model_version14$coefficients[4]
sd(rates$pctPrivateHC) * model_version14$coefficients[5]
sd(rates$pctEmployerHC) * model_version14$coefficients[6]
sd(rates$pctMarried) * model_version14$coefficiends[7]
sd(rates$pctMarriedHouse) * model_version14$coefficients[8]
sd(rates$avgHouse) * model_version14$coefficients[9]
sd(log(rates$population)) * model_version14$coefficients[13]

```




```{r}
quadData <- data.frame(pctPoverty = seq(min(rates$pctPoverty), max(rates$pctPoverty), by = 0.01))
quadData$pctBach <- mean(rates$pctBach)
quadData$pctUnemployed <- mean(rates$pctUnemployed)
quadData$pctPublicHC <- mean(rates$pctPublicHC)
quadData$pctPrivateHC <- mean(rates$pctPrivateHC)
quadData$pctEmployerHC <- mean(rates$pctEmployerHC)
quadData$pctMarried <- mean(rates$pctMarried)
quadData$pctMarriedHouse <- mean(rates$pctMarriedHouse)
quadData$avgHouse <- mean(rates$avgHouse)
quadData$population <- median(rates$population)
quadData$region <- "Southeast"
quadData$Fitted <- predict(model_version12, newdata = quadData)
quadData$Fitted1 <- predict(model_version13, newdata = quadData)
quadData$Fitted2 <- predict(model_version14, newdata = quadData)

plot(rates$medianIncome ~ rates$pctPoverty)
lines(quadData$pctPoverty, quadData$Fitted, col = "blue")
lines(quadData$pctPoverty, quadData$Fitted1, col = "red")
lines(quadData$pctPoverty, quadData$Fitted2, col = "green")
```


```{r}
model_version15 <- lm(medianIncome ~ pctBach + pctUnemployed + pctPublicHC + pctPrivateHC + pctEmployerHC + pctMarriedHouse + avgHouse + pctPoverty + I(pctPoverty^2) + log(population) + region, data = rates)
summary(model_version12)
summary(model_version15)
```

```{r}
vif(model_version15)
```


```{r}
summary(rates)
summary(model_version15)
```





```{r}
sd(rates$pctBach) * model_version15$coefficients[2]
sd(rates$pctUnemployed) * model_version15$coefficients[3]
sd(rates$pctPublicHC) * model_version15$coefficients[4]
sd(rates$pctPrivateHC) * model_version15$coefficients[5]
sd(rates$pctEmployerHC) * model_version15$coefficients[6]
sd(rates$pctMarriedHouse) * model_version15$coefficients[7]
sd(rates$avgHouse) * model_version15$coefficients[8]
sd(log(rates$population)) * model_version15$coefficients[11]

```



